Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan

### 2020-06-27 analysis/paper/paper.Rmd
finished small edits based on R1's comments

@@

Ceramics analyzed in this paper come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)). The chronology of the archaeological deposits at Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007]. This paper focuses on the upper component, dated from AD 1350 to AD 1850, covering the late Iron Age and the historical period, which we define as the European occupation in Taiwan in the early 17th century. The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 [@Andrade2007]. In 1642, the Spanish was expelled by the Dutch in northern Taiwan, who took over their forts at Helping Dau in Keelung, and in Tamsui. Since then, western Taiwan was mostly under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].

```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text. Map data from naturalearthdata.com", fig.width=image_width_2_col_inch}
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.jpg"))
```

@@

The archaeological record of Kiwulan's upper component shows traces of foreign contacts, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century. Imported ceramics from mainland China, stoneware, and ornaments such as beads have been found in the upper component, indicating frequent long-distance trade activities with Europeans and Chinese merchants. In addition to artifacts, archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007]. To compare different foreign influences, we classified the upper component into three chronological phases: the pre-European, European, and Chinese periods. We identified these phases using 32 previously published radiocarbon ages [@Chen2007], excavation depth measurements, stratigraphic details reported by the excavators (color, texture, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011]. The deposit shows signs of continuous human occupation during each of the three phases across the sampling area.

The most abundant artifacts in the upper component are locally made ceramics, which are distributed throughout the sequence and across the site. More than 550,000 sherds were excavated, and around 1,200 vessels could be partially reconstructed (i.e. complete rim, or complete base). There are only two forms of local vessels, a cooking pot and a steamer made of two cooking pots stacked together with a clay filter layer in between. Those pots forms show high consistency in shape. They have a globular body with a short neck and wide mouth (Figure \@ref(fig:KWL-pots)). The exterior surface below the neck is decorated with a wide variety of impressed geometric motifs. These vessels were probably used for cooking, with evidence of charred residues and carbon deposits frequently observed on vessel interiors and soot on their exteriors. They are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand. This kind of pot has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain.

Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite (15-40%), metasandstone (1-10%),  sandstone fragments (1-6%), quartz (1-5%), and trace amounts of feldspar and slate. The size of the particles ranges from 500 micron to 1300 micron. In general, the globular vessel fabric presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz. The composition matches the mineralogical composition of local raw materials in the Yilan Plain [@Chen2016]. There are no substantial changes in the inclusions over time, indicating continuity in pottery fabric composition across the three periods.

@@

We used all `r total_no` vessels that could be reconstructed with rim, body and base parts and securely provenanced to pre-contact contexts (n = `r pre_no`), post-European contexts (n = `r Euro_no`), and Chinese contact contexts (n = `r Chi_no`) for elliptic Fourier analysis.

@@

We used scans of pottery drawings (Figure \@ref(fig:KWL-pots)) acquired from the Bureau of Cultural Affairs in Yilan. All drawings present a two-dimensional view of vessel cross-sections with indications of metric measurements. The scanned drawings were imported into Inkscape (http://inkscape.org) where outlines were traced manually. Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel. Geometric morphometric analyses were conducted using the R software (R Core Team, 2019) and the functions included in the Momocs, a R package intended to quantify and analyze shapes [@Bonhomme2014]. The digilized outlines were converted into a list of successive xy pixel coordinates for elliptic Fourier analysis (EFA). The harmonic coefficients generated by EFA were analyzed by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation.

@@

We found that 13 harmonics captured 99% of the total harmonic power in our elliptic Fourier coefficients of `r total_no` vessels from three phases. Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot) shows vessel shape changes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods, with the most greatest differences between pre-European and Chinese periods.

@@

The first two principal components (PCs) of the PCA on the elliptic Fourier coefficients explain `r vari_two`% of the total variance, of which `r pc1`% is explained by the first principle component. With the third component, the first three principal components explain `r vari_three`% of the total variance. PC1 captures the height of the vessels, from tall to short, and the roundness of the body from round to oval-shaped (Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot)). PC2 relates to the neck and mouth constriction, from narrow to wide. PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare of the neck, from a curved to straight shape. The results reflect a large overlap in shapes from three occupations phases, especially for shapes in the pre-European and post-European periods. However, the spread of shape distribution indicates a wider variation in shapes in the pre-European and post-European periods compared to those in the Chinese period along both PC1 and PC2 axes. In other words, we find a decrease in shape variance in the Chinese period that is evident in the shorter height and narrower mouth of vessels used in that period.

@@

To test for differences in the distributions of shape variables indicated by the PC scores shown in Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot), we used a multivariate analysis of variance (MANOVA) test to compare pairwise combinations across our three occupation phases. Table \@ref(tab:MANOVA-table) shows significant differences in shape between Pre-European and Post-European phases (p = $`r pval_post_pre`$), and Pre-European and Chinese phases (p = $`r pval_chi_pre`$). These results match the differences in the visualization of average shapes between the occupations in the thin plate splines (Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot), see left). Although there is considerable overlap of shape variables between the Pre-European and Post-European phases, their PC scores are significantly different from each other. There is little difference between vessel shapes in the Post-European and Chinese contact periods.

@@

Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans [@Cheng2008; @Wang2011], hinting at an emergence of social inequality in the Indigenous community. We have examined ceramic vessel shape standardization to measure craft specialization as a proxy for social differentiation [@Costin2001; @Junker1999]. The result of our MANOVA test shows there is a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods. The average shape presents a round body with a wider rim and neck before European contact, and turns to a more oval-shaped body with narrower rim and neck after the European presence. These changes even more pronounced during the Chinese contact period. In general, pottery vessels become shorter in height over time, leading to an oval-shaped body.

@@

Our results offer tentative support for the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society. If increased shape standardization is a reliable indicator of craft specialization, then we may be seeing evidence of a shift from corporate (group-based, distributed, collective, cooperative) to network (individual-based, competitive) organization [@Blanton1996; @Feinman1995; @Feinman2000; @Feinman2000political; @Feinman2010]. However, strong claims for an emergence of social complexity resulting from foreign contact at Kiwulan will need support from multiple and diverse sources of evidence that are beyond the scope of this paper. We find vessel shapes were more standardized in the Chinese period than the European period.

@@

We recognize that the shape variation reported here is subtle, and this invites consideration of the possibility that the absence of major changes in vessel shape at Kiwulan may have been an act of resistance to foreign influence. Continuities in vessel shape over time draws our attention to the endurance of traditional pottery production practices amid intrusions from Europeans and Chinese. In a culture contact situation, social identity may be expressed through material practices to show cultural homogeneity and distinction from other groups [@Voss2005]. It is also important to recognize that social identity might be more complicated in a colonial context, and may be more than a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008]. Shamaoshan cemetery (3BC- 4AD) in Southwest China shows that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019]. A similar dynamic may have occurred at Kiwulan, with vessel shape indicating both acceptance of foreign influence through increased shape standardization, and resistance through the overall continuity in vessel shape. Vessel shape can be viewed as a symbol as an expression the Indigenous identity and social boundaries because shape is a highly visible trait compared with other features of pottery [cf. @Roux2015]. Although there is an increase in number of imported ceramics over time at Kiwulan, the production of the local pottery continued and became more standardized. This might imply not only a utilitarian function of the local vessels, but a deliberate action to emphasize the local pottery tradition as their cultural custom as foreign contact intensified.

@@

This study demonstrates an application of elliptic Fourier analysis on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences. Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust way to identify and statistically assess differences in shape standardization. The direct relationship between foreign influences and standardization of ceramic shape was tested on pottery from Kiwulan, a large Iron Age Indigenous settlement in northeastern Taiwan. Lower variation in ceramic shape was identified after European presence began, and even lower variation during the period of Chinese presence. Our findings help to understand the factors that may lead to standardization of pottery production in a pericolonial interaction context. More homogeneous shapes during the contact periods, without any changes in clay paste composition or production technique, suggests that shape standardization was intentional. The results here further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign contact in pericolonial contexts. Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.

>>>>

<!--discussion of standardization in material culture using GM approaches-->

@@

Ceramics analyzed in this study come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)). The chronology of the archaeological deposits consist of two cultural components, the upper and the lower, with a sterile layer in between [@Chen2007]. We focus on the upper component, dated from AD 1350 to 1850, spans the late Iron Age and the historical period, of which the start is defined as the European occupation in Taiwan in the early 17th century. The Dutch first occupied southern Taiwan in 1624, followed by the Spanish in northern Taiwan in 1626 [@Andrade2007]. In 1642, the Spanish was expelled by the Dutch, who then occupied the previous Spanish forts at Helping Dau in Keelung, and in Tamsui. Western Taiwan remained under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].

```{r, KWL-map, fig.cap = "Map illustrating the location of Kiwulan, and other locations in northern Taiwan that are named in the text. Map data is from naturalearthdata.com", fig.width=image_width_2_col_inch}
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.jpg"))
```

@@

The archaeological record of Kiwulan's upper component shows traces of foreign contact, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century. Imported ceramics from mainland China, stoneware, and ornaments such as beads have been recovered in the upper component, indicating frequent long-distance trade activities with Europeans and Chinese merchants. Archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007]. To compare different foreign influences, we classified the upper component into three chronological phases: pre-European, European, and Chinese. We identified these phases based on 32 previously published radiocarbon ages [@Chen2007], excavation depth measurements, stratigraphic details reported by the excavators (color, texture, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles employed by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011]. The deposit exhibited signs of continuous human occupation in each of the three phases with no apparent breaks.

The most abundant artifacts in the upper component are locally manufactured ceramics, which are distributed throughout the temporal sequence, and across the study site. More than 550,000 sherds were excavated, and around 1,200 vessels could be partially reconstructed (i.e. complete rim or base). There are two shapes of locally-manufactured vessels; a cooking pot and a steamer made of two cooking pots stacked together with a clay filter between. Those vessel shapes demonstrate suites of standard morphological components. Each has a globular body with a short neck and wide mouth (Figure \@ref(fig:KWL-pots)). The exterior surface below the neck is decorated with a variety of impressed geometric motifs. These vessels were potentially used for cooking indicated by evidence of charred residues and carbon deposits frequently observed on vessel interiors, and soot on vessel exteriors. Firing resulted in orange and brownish color with a fully oxidized core, or a reduced core with oxidized fringes [@Chen2007]. Finger impressions and seams, usually on the interior, indicate that the vessels were shaped by hand, and that the seems between the slabs were pinched. This kind of vessel has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain [@NMTH2005].

Petrographic analysis for 34 thin sections presents a high percentage of inclusions (15-50%), including argillite (15-40%), metasandstone (1-10%),  sandstone fragments (1-6%), quartz (1-5%), and trace amounts of feldspar and slate. Particle sizes of range from 500 to 1300 microns. In general, the globular vessel fabric presents a mixture of fine, rounded argillite with a small amounts of rounded metasandstone and rounded to sub-angular monocrystalline quartz. This composition is consistent with the mineralogical composition of local raw materials found in the Yilan Plain [@Chen2016]. There are no substantial changes in the inclusions over time, indicating continuity in pottery fabric composition across the three periods. <!--Was this tested using statistics? If so, then state that there are no significant changes in the inclusions over time, and include that script in your supplementary materials.-->

@@

The sample consists of `r total_no` reconstructed vessels with rim, body and base parts, and we securely provenanced to pre-contact (n = `r pre_no`), post-European (n = `r Euro_no`), and Chinese contact contexts (n = `r Chi_no`).

@@

We used scans of pottery drawings acquired from the Bureau of Cultural Affairs in Yilan (Figure \@ref(fig:KWL-pots)). All drawings provide a two-dimensional view of vessel cross-sections, and metric measurements. The scanned drawings were imported into Inkscape (http://inkscape.org) for digitization where outlines were manually traced. In those instances where only one side of the cross-section image was available, these were mirrored and joined to create a closed outline for each vessel. Analyses were conducted in R software (R Core Team, 2019) using functions included in the Momocs package for quantifying and analyzing shapes [@Bonhomme2014]. Outlines were converted into a list of successive x, y pixel coordinates for EFA. We analyzed harmonic coefficients by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation.

@@

We found that 13 harmonics captured 99% of the total harmonic power in the elliptic Fourier coefficients of `r total_no` vessels from three phases. Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot) illustrates significant differences in vessel shapes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods, with the greatest differences between pre-European and Chinese periods.

@@

The first two principal components (PCs) of the PCA on the elliptic Fourier coefficients explain `r vari_two`% of the total variance, of which `r pc1`% is explained by the first principle component. With the third component, the first three principal components explain `r vari_three`% of the total variance. PC1 captures the height of the vessels, from tall to short, and the roundness of the body from round to oval-shaped (Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot)). PC2 relates to the neck and mouth constriction, from narrow to wide. PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare in the neck, from a curved to straight shape. The results reflect a large overlap in shapes from three occupations phases, especially for shapes in the pre-European and post-European periods. However, the spread of shape distribution indicates a wider variation in shapes in the pre-European and post-European periods compared to those in the Chinese period along both PC1 and PC2 axes. In other words, we find a decrease in shape variance in the Chinese period that is evident in the shorter height and narrower mouth of vessels used in that period.

@@

To test for differences in the distributions of shape variables indicated by the PC scores shown in Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot), we used a multivariate analysis of variance (MANOVA) test to compare pairwise combinations across the three occupation phases. Table \@ref(tab:MANOVA-table) demonstrated the significant differences in shape between Pre-European and Post-European phases (p = $`r pval_post_pre`$), and Pre-European and Chinese phases (p = $`r pval_chi_pre`$). These results are consistent with the differences in the visualization of average shapes between the occupations in the thin plate splines (Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot), see left). Although there is considerable overlap of shape variables between the Pre-European and Post-European phases, their PC scores differ significantly. There is not a significant difference in vessel shapes between the Post-European and Chinese contact periods.

@@

Previous investigations at Kiwulan suggested an unequal distribution of prestige goods, trade ornaments in particular, following the appearance of Europeans [@Cheng2008; @Wang2011], which hinted at the emergence of social inequality within the Indigenous community. Ceramic vessel shape standardization was examined to measure craft specialization as a proxy for social differentiation [@Costin2001; @Junker1999]. The result of our MANOVA demonstrates a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods. The average shape presents as a round body with a wide rim and neck before European contact, which shifts to a more oval-shaped body with narrower rim and neck after the European presence. Shape changes are more pronounced during the Chinese contact period. In general, vessels become shorter in height over time, leading to an oval-shaped body.

@@

Our results offer tentative support for the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society. If increased shape standardization is a reliable indicator of craft specialization, then we may be seeing evidence of a shift from corporate (group-based, distributed, collective, cooperative) to network (individual-based, competitive) organization [@Blanton1996; @Feinman1995; @Feinman2000; @Feinman2000political; @Feinman2010]. However, strong claims for an emergence of social complexity resulting from foreign contact at Kiwulan will need support from multiple and diverse sources of evidence that are beyond the scope of this paper. We find that vessel shapes were more standardized during the Chinese period than the European period.

@@

The shape variation reported here is subtle, and invites consideration of the possibility that the absence of major changes in vessel shape at Kiwulan may have been an act of resistance to foreign influence. Continuities in vessel shape over time draws our attention to the endurance of traditional pottery production practices amid intrusions from Europeans and Chinese. In a culture contact situation, social identity may be expressed through material practices as a means of expressing cultural homogeneity and distinction from other groups [@Voss2005]. It is also important to recognize that social identity might be more complicated in a colonial context, and may be representative of more than a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008]. Shamaoshan cemetery dating from BC 3 to AD 4 in Southwest China suggests that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019]. A similar dynamic may have occurred at Kiwulan, with vessel shape indicating both acceptance of foreign influence through increased shape standardization, and resistance through the overall continuity in vessel shape. Vessel shape can be viewed as a symbolic expression the Indigenous identity and social boundaries because shape is a highly visible trait compared with other features of pottery [cf. @Roux2015]. Although there is an increase in number of imported ceramics through time at Kiwulan, production was continuous, and increasingly standardized. This might imply not only the utilitarian function, but a deliberate and increased emphasis on the local ceramic tradition, their cultural custom, as a response to intensified foreign contact [cf. @Acabado2017].

@@

This study demonstrated an application of EFA on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences. Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust method of identifying and statistically assessing differences in shape standardization. The direct relationship between foreign influences and standardization of ceramic shape was tested on ceramics from Kiwulan, a large Iron Age Indigenous settlement in northeastern Taiwan. Lower variation in ceramic shape was identified after European presence began, and even lower variation during the period of Chinese presence. Our findings help to expand upon those factors that may lead to the standardization of ceramic production in a pericolonial interaction context. More homogeneous shapes during the contact periods, without any changes in clay paste composition or production technique, suggest that shape standardization was intentional. The results further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign contact in pericolonial contexts. Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.
